---
title: "R Basics"
author: "Madison Frank"
date: "October 2018"
output:
  html_document:
    df_print: paged
    code_folding: show
    theme: lumen
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting Started 

**This document provides a very brief overview to basic progamming in R adapted from training provided by Prof. Ross Ihaka.** 

Further information can be found at https://www.stat.auckland.ac.nz/~ihaka/?Teaching

*Topics include:* 

* Setting Up
* R 101: The Basics
* Creating Plots
* Writing Functions
* R and Hilltop  
* Report Writing

## Set up 

To start we need to download R, as of Oct 2018 the latest version is 3.5.1 (Feather Spray). I also highly recommend downloading RStudio - an integrated development environment (IDE) for R.

  <div align="center">[R](https://www.r-project.org/)  |  [RStudio](https://www.rstudio.com/)</div>
  

# Starting to Code
## Basic calculations 
One way to think of R is as a very powerful calculator. If we type a calculation into the console an answer is printed out.

```{r} 
# Performing calculations using mathematical notation
2 + 2

# Precedence can be set using parentheses
(10 + 5) / 2
```

Once R has completed the computation the answer is discarded. To 'save' a result you can assign a name using <- or = to create a _variable_. Variables can then be used in subsequent expressions.

```{r}
# Create a variable by assigning a name 'a'
a <- 5^2
```
Notice no result is printed

```{r}
# Print the result named 'a'
a

# Calculate the square root of 'a'
sqrt(a)
```

## Vectors 
R operates on named data structures - the most basic of which is a vector. Vectors contain an indexed collection of values of the *same type*.

Vectors are usually created with the function c()
```{r}
x <- c(1, 3, 5, 7, 9)
y <- c(TRUE, FALSE, TRUE)
```

Check the type of a variable 
```{r}
class(x)
class(y)
```

Determine the number of elements in a vector
```{r}
length(x)
length(y)
```

```{r}
# Be careful of mixing types
c(1, 5.5, 7)

c(1, TRUE, "Hello")

```

## Number sequences 
Creating vectors by a sequence
```{r}
5:10
seq(1, 10, by = 2)
-12.5:12
```

## Vector arithmetic 
```{r}
x
x + 10
sqrt(x)
1:10/2
```

## Subsetting vectors 
Values can be subset out a vector using square brackets [ ]

```{r}
x

x[3]

x[2:4]

x[-3]
```

## Conditional subsetting 

```{r}
x > 5
x[x > 5]
```

## Named vectors 
```{r}
names(x) <- c("Jan", "Mar", "May", "Jul", "Sep")
x

x["May"]
```

## Summarizing data 
```{r}
c(mean = mean(x), sd = sd(x))
quantile(x)
cumsum(x)
```

**Tip:** To get information about a function try **?_function name_**

## Dataframes 
Dataframes are one of the ways R presents tabular data. 

A dataframe is comprised of columns and rows, both of which can be named, of the same length. Each column holds the same type of data.

```{r}
rain <- read.csv("Data/MIL_DailyRainfall.csv")
rain$Date <- as.Date(rain$Date)

# Display the start or end of a dataframe with head() and tail()
head(rain)
```

## Subsetting dataframes 
Dataframes can be subset using [row, column]
```{r}
rain[1, 1]

rain[1,]

head(rain[,1])
```

## Building a PCAL dataframe 

```{r}  
# Create a month column for calculating statistics
rain$Month <- factor(format(rain$Date, "%b"), levels = month.abb)

head(rain)
```

Note that you can call a column by name with $

```{r}
# # Calculate monthly total rainfall
# rainTotal <- tapply(rain$Rainfall, rain$Month, sum)
# 
# # Create a dataframe with month and total rainfall
# df <- data.frame(Month = month.abb,
#                  Total = rainTotal)
# 
# head(df)
```

### Add min, mean and max to the dataframe
```{r}
# Add min, mean and max to the dataframe
pcal <- cbind(df, 
              Min = tapply(rain$Rainfall, rain$Month, min), 
              Mean = tapply(rain$Rainfall, rain$Month, mean), 
              Max = tapply(rain$Rainfall, rain$Month, max))

# Drop the rownames
row.names(pcal) <- 1:12


head(pcal)
```

### Rounding numbers to make the table more user friendly
```{r}
# Add min, mean and max to the dataframe
pcal <- cbind(df, 
              Min = tapply(rain$Rainfall, rain$Month, min), 
              ### <b>
              Mean = round(tapply(rain$Rainfall, rain$Month, mean), 2), 
              ### </b>
              Max = tapply(rain$Rainfall, rain$Month, max))

# Drop the rownames
row.names(pcal) <- 1:12


head(pcal)
```

# Making plots 
## Base R plotting 
R has built in plotting capabilities - to view more details about possible graphical parameters try ?plot

```{r}
x <- data.frame(Month = 1:12, 
                Flow = c(25.4, 38.6, 27.3, 56.5, 53.4, 48.7, 
                         59.1, 62.1, 60.1, 69.8, 25.3, 13.5))
```

## A basic plot 

```{r}
plot(x)
```

## Improving the plot 

```{r}
plot(x, main = "Manawatu at Teachers College", pch = 16, col = 1:12, type = "b",
     xlab = "2017", ylab = "Avg Monthly Flow (cumecs)")
```

## Adding more information to our plot

```{r}
# Choosing colours for the points based on flow values
cols <- vector(length = 12)
cols[x$Flow < 30] <- "red"
cols[x$Flow > 60] <- "blue"
cols[x$Flow >= 30 & x$Flow <= 60] <- "black"
```

----_

```{r plot1, eval = FALSE}
plot(x, main = "Manawatu at Teachers College Monthly Flow", xaxt = "n",
     xlab = "2017", ylab = "Avg Monthly Flow (cumecs)")
rect(6, 0, 9, 72, col = "lightgrey")
text(7.5, 45, "Winter")
lines(x$Month, x$Flow)
points(x, pch = 16, col = cols)
abline(h = 30, lty = 2)
abline(h = 60, lty = 2)
text(c(0, 20), "Low flows")
text(c(0, 65), "High flows")
axis(1, x$Month, month.abb)
```

----

```{r plot1, echo=FALSE, eval=TRUE}
```

## Add a legend

```{r legendPlot}
par(oma=c(0, 0, 0, 5))

plot(x, main = "Manawatu at Teachers College Monthly Flow", xaxt = "n",
     xlab = "2017", ylab = "Avg Monthly Flow (cumecs)")
rect(6, 0, 9, 72, col = "lightgrey")
text(7.5, 45, "Winter")
lines(x$Month, x$Flow)
points(x, pch = 16, col = cols)
abline(h = 30, lty = 2)
abline(h = 60, lty = 2)
text(c(0, 20), "Low flows")
text(c(0, 65), "High flows")
axis(1, x$Month, month.abb)

legend(x = par('usr')[2], y = par('usr')[4], bty='n', xpd=NA, pch = 16, col = unique(cols),
       legend = c("Low flow", "Moderate flow", "High flow"))
```

----

```{r legendPlot, echo=FALSE, eval=TRUE}
```

## Other libraries for plotting
There are a multitude of different packages for plotting. Some of the most popular include:

* ggplot2 
* htmlwidgets
  
  ----

### ggplot2

ggplot2 is part of the tidyverse package collection. Very basically it works by mapping the aesthetics of a dataset and then adding layers, scales, facets, etc... to build up a plot. 
  If you would like to learn more about tidyverse coding I highly recommend [R for data science](https://r4ds.had.co.nz/index.html) by Hadley Wickham.
  
```{r ggplot}
library(ggplot2)

ggplot(data = mpg) + 
geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

  htmlwidgets are a collection of packages which provide an R interface to JavaScript libraries to create interactive visualisations. There are over 50 widgets available including dedicated options for mapping and timeseries. 
  To see what widgets exist visit [**htmlwidgets** for R](https://www.htmlwidgets.org/index.html)

----

### htmlwidgets

```{r plotly, message=FALSE}
library(plotly)
plot_ly(data = iris, x = ~Sepal.Length, y = ~Petal.Length, color = ~Species, type = "scatter")
```

----

```{r dygraph}
library(dygraphs)
dygraph(nhtemp, main = "New Haven Temperatures") %>% 
  dyRangeSelector(dateWindow = c("1920-01-01", "1960-01-01"))
```

----

```{r leaflet}
library(leaflet)
leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addMarkers(lng=174.768, lat=-36.852, popup="The birthplace of R")
```

# Functions in R
## Built in functions 
Functions in R are denoted as: _function name_( )

Function parameters or arguments are separated by a comma and may be named.  

We have used a number of functions already such as; 
```{r}
x <- c(1:5)

length(x)

plot(x, main = "A Plot", col = x, pch = 16)
```

**Tip:** To view a function's arguments try typing in _function name_() and hit tab while the cursor is _inside_ the parentheses.

## Vectorization 
Vectorization is a fundamental principle in R. It refers to the ability to perform operations on a vector as a whole rather than element by element.

For example we can perform vector arithmetic;
```{r}
x <- c(1, 2, 3)
y <- c(10, 20, 30)
x+y
```

Or we can pass a vector to a function and receive a result for each element;
```{r}
sqrt(x)
```

## Recycling 
The recycling rule is how R handles vector operations on vectors of _different_ length. When the operation requires that both vectors be the same length R will automatically recycle the values of the shorter vector until both vectors are the same length.

For example;
```{r}
x + 10
x[c(TRUE, FALSE)]
```

## A Calculator Curiosity
This example is taken from the book _Professor Stewart's Hoard of Mathematical Treasures_ by Ian Stewart. The solution code was provided by Prof. Ross Ihaka.

(8 x 8) + 13

(8 x 88) + 13

(8 x 888) + 13

(8 x 8888) + 13

(8 x 88888) + 13

(8 x 888888) + 13

(8 x 8888888) + 13

(8 x 88888888) + 13

## A Calculator Curiosity: A vectorised solution 
The sequences of 8 can be decomposed into 8, 8 + 80, 8 + 80 + 800, ect...

Those values are the cummulative sum of the sequence 8, 80, 800, etc...

That sequence is equivalent to 8x10^0, 8x10^1, 8x10^2, etc...

So our code is as follows;
```{r}
n = 8
cumsum(8 * 10^(1:n - 1))
8 * cumsum(8 * 10^(1:n - 1)) + 13
```

## Control flow 
There are several control constructs available in R, most commonly; 

```{r, eval=FALSE}
if(condition){
  expression
}
```

```{r, eval=FALSE}
if(condition){
  expression_1
}else{
  expression_2
}
```

```{r, eval=FALSE}  
for(variable in vector){
  expression
}
```

## A Calculator Curiosity: A control flow solution 
How can we solve the calculator curiosity problem using control flow?

The most logical method for solving our problem is using a for loop.


Again building up the 8, 88, 888... is key. 

We can do this by multiplying the previous value by 10 and adding 8.

## A Calculator Curiosity: A control flow solution 

```{r}
numSeq = 8

for (i in 1:8) {
  print(8 * numSeq + 13)
  
  numSeq <- 10 * numSeq + 8
}
```

## Writing your own functions 
At some point you may want to start writing your own functions. 

This can easily be acomplished in R using the structure

```{r, eval=FALSE}
functionName <- function(arg_1, arg_2 = defaultValue){

                         Expressions to do something

                         return(results)
                }
```

You can then call your function with the default value

_functionName_(x)

or if you would like to overwrite the default value

_functionName_(x, arg_2 = y)

## A Calculator Curiosity: A function 
Now that we can calculate the problem, how can we make a function to output _any_ number of answers?

Our vectorised version is already a simple expression using 1 input:
```{r}
curiosity_1 <- function(n) 8 * cumsum(8 * 10^(1:n - 1)) + 13
```

```{r}
curiosity_1(5)
```

## A Calculator Curiosity: A function 

Our control flow solution is a bit more complicated. We need some way to store the results instead of just printing them out.

```{r}
curiosity_2 <- function(n){
  ans <- numeric(n)
  numSeq = 8

  for (i in 1:n) {
    ans[i] <- 8 * numSeq + 13
    
    numSeq <- 10 * numSeq + 8
  }
  return(ans)
}
```

```{r}
curiosity_2(3)
```


